Computational methods in landscape ecology

### 2024-09-07 Manuscript.qmd
Adding genetics section by ADF

author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affil-id: 1
    attributes:
      - corresponding: boolean
  - name: Jakub Nowosad
    orcid: 0000-0002-1057-3721
    email: nowosad.jakub@gmail.com
    affil-id: 2,3
  - name: Marti Bosch
    orcid: 0000-0001-8735-9144
    email: marti.bosch@epfl.ch
    affil-id: 4
  - name: Martin Jung
    orcid: 0000-0002-7569-1390
    email: jung@iiasa.ac.at
    affil-id: 1

affiliations: 
  - id: 1
    name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
  - id: 2
    name: Adam Mickiewicz University, Poznan, Poland
  - id: 3
    name: University of Münster, Münster, Germany
  - id: 4
    name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland

>>>>

author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affil-id: 1
    attributes:
      - corresponding: boolean
  - name: Jakub Nowosad
    orcid: 0000-0002-1057-3721
    email: nowosad.jakub@gmail.com
    affil-id: 2,3
  - name: Martí Bosch
    orcid: 0000-0001-8735-9144
    email: marti.bosch@epfl.ch
    affil-id: 4
  - name: Martin Jung
    orcid: 0000-0002-7569-1390
    email: jung@iiasa.ac.at
    affil-id: 1
  - name: Alida de Flamingh
    orcid: 0000-0003-1223-6654
    email: adeflamingh@gmail.com
    affil-id: 5

affiliations: 
  - id: 1
    name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
  - id: 2
    name: Adam Mickiewicz University, Poznan, Poland
  - id: 3
    name: University of Münster, Münster, Germany
  - id: 4
    name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland
  - id: 5
    name: Carl R. Woese Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA

### 2024-09-07 Sections/03-connectivity.qmd
Adding genetics section by ADF

Resistance surfaces are one of the most popular approaches to model connectivity and developed from simple isolation-by-distance approaches. 
The approach uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012EstimatingLandscape].
The latest developments of resistance surfaces have recently been reviewed [@Dutta2022OverviewComputational].
The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003ApplicationLeastcost; @Etherington2016LeastcostModelling] or circuit theory [@McRae2008UsingCircuit].
However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].

>>>>

Resistance surfaces are one of the most popular approaches to model connectivity and developed from simple isolation-by-distance approaches. 
The approach uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012EstimatingLandscape].
The latest developments of resistance surfaces have recently been reviewed [@Dutta2022OverviewComputational].
The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003ApplicationLeastcost; @Etherington2016LeastcostModelling] or circuit theory [@McRae2008UsingCircuit].
However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
Software to model connectivity based on resistance surfaces include *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].

### 2024-09-07 Sections/05-simulations.qmd
Adding genetics section by ADF

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024LeastcostNetwork].
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization [@VanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand; @Cambui2015GradientLandSoftware]

@@

Digital Twins include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @DeKoning2023DigitalTwins] and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for ecology [@DeKoning2023DigitalTwins].
Digital twins are increasingly applied in agricultural landscapes research, including topics such as crop and land cultivation, livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows them to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
Thus, Digital Twins can provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@DeKoning2023DigitalTwins].

>>>>

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024LeastcostNetwork].
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand; @Cambui2015GradientLandSoftware]

@@

Digital Twins include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins] and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for ecology [@deKoning2023DigitalTwins].
Digital twins are increasingly applied in agricultural landscapes research, including topics such as crop and land cultivation, livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows them to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
Thus, Digital Twins can provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@deKoning2023DigitalTwins].

### 2024-09-07 Sections/07-genetics.qmd
Move software to corresponding sections
Adding genetics section by ADF

<!-- Person responsible: De Flamingh -->

# Landscape genetics

>>>>

<!-- Person responsible: de Flamingh -->

# Landscape genetics {#sec-genetics}

Genetic data can be applied in complement with ecological data to integrate evolutionary processes and patterns with the ecological processes and patterns typically studied in landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics]. 
The field of landscape genetics unites molecular population genetics, spatial statistics, and landscape ecology, and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics]. 
Landscape genetic approaches have traditionally been used to inform landscape ecology applications, e.g., through connectivity modeling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
Using holistic approaches that consider evolutionary processes and patterns in addition to only ecological data can fortify results, for example, by improving predictive models of species range shifts in response to climate change [e.g., @Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [e.g., @Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], by allowing for the interpretation of spatial structuring in context of socio-cultural connections [e.g., @Henson2021ConvergentGeographic], and by understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].

Advances in molecular methodologies and the synergistic development of bioinformatic and computational approaches to analyze large-scale genomic data have now enabled the integration of genome-wide data with spatial ecological data. 
This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers may apply genomic data across spatiotemporal scales, including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape] or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics]. 
Analytical frameworks such as the FOLDS model [gene flow, genetic offsets, genetic load, dispersal and SDMs, @Aguirre-Liguori2021EvolutionaryGenomics], constrained coordination where covarying sets of genotypes are correlated with multivariate environments through redundancy analysis [RDA, @Capblancq2021RedundancyAnalysis] and spatial simulations provide promising potential for future landscape genomic studies. 
Such studies may explore interactions between complex evolutionary processes, e.g., demography, multidimensional (e.g., strength and direction) gene flow and migration across spatiotemporal scales, genetic differentiation across landscapes, genomic load and adaptive potential, and the interpretation of these evolutionary processes in context of landscape ecology [@Balkenhol2017LandscapeGenomics; @House2018EvaluatingMethods; @Fenderson2020SpatiotemporalLandscape; @Marcus2021FastFlexible; @Chevy2024PopulationGenetics].
